Supermartingales and Termination Analysis of Statistical PCF

### 2021-04-26 LICS21/ranking.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

\begin{example}
\label{ex:tY finite does not imply t finite}
The term $M = \Xi \, (\lambda x. x+1)$ where
\[
\Xi := \tY \lambda f \, n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n \, x))}
\] 
terminates with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
\end{example}

@@

\begin{example}[Geometric distribution]
% Consider the term
% \begin{align*}
% &(\tY \lambda f, n. \\
% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
% &) \, \underline{0}
% \end{align*}
%\[
%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
%\big) \, \underline{0}
%\]
Let
\[
\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
\]
The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
Despite its simplicity, its $\mathit{Rch}$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
%\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK. Thanks.} 
The meanings of the substitutions (to be applied to the contractum) and conditions that label some of the reductions in the diagram should be self-explanatory.

@@

The definition of rankability is also inconvenient for syntactic sugar. It could be useful, for example, to define $M \oplus_p N := \tif{\tsample - p < 0} M N$, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$. Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.

>>>>

\begin{example}
\label{ex:tY finite does not imply t finite}
The term $M = \Xi \, (\lambda x. x+1)$ where
\[
\Xi := \tY \lambda f \, n. \tif{\tsample - 0.5 < 0}{n \, \underline 0}{f (\lambda x. n (n \, x))}
\] 
is $\tY$-PAST and rankable, terminating with only 2 $\tY$-reductions on average, i.e.,~$\expect{T_M^\tY} = 2$, but applies the increment function $2^n$ times with probability $2^{-n-1}$ for $n \geq 0$, which diverges, i.e.,~$\expect{T_M} = \infty$.
\end{example}

@@

\begin{example}[Geometric distribution]
% Consider the term
% \begin{align*}
% &(\tY \lambda f, n. \\
% &\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
% &) \, \underline{0}
% \end{align*}
%\[
%\big(\tY \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
%\big) \, \underline{0}
%\]
Let
\[
\Theta := \lambda f, n. \tif{\tsample - 0.5 < 0}{n}{f (n+1)}. 
\]
The term  $(\tY \, \Theta) \, \underline 0$ generates a geometric distribution.
Despite its simplicity, its $\mathit{Rch}$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.
%\lo{TODO: reduce the figure to avoid overflow.}\akr{It still overflows the gutter, but not the actual text of the other column.}\lo{OK. Thanks.}

@@

The definition of rankability is also inconvenient for syntactic sugar. It could be useful, for example, to define \mbox{$M \oplus_p N := \tif{\tsample - p < 0} M N$}, where $M \oplus_p N$ reduces to $M$ or $N$, depending on the first value of $s \in \entrosp$, with probability $p$ resp.~$(1-p)$. Technically though, it reduces first to $\tif{\underline r - p < 0} M N$ for all $r \in I$, so those terms all need values of the ranking function too.

### 2021-04-26 LICS21/related.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
Note that $n$-PHORS is strictly less expressive then order-$n$ call-by-name PPCF:
% (because the underlying recursion schemes are not Turing complete):
the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
%\refExample{complexExample} 
are not definable as PHORS.

@@

The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee Church-Rosserness, but is also a little more restrictive in some cases than is necessary for this purpose. For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples. Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine. A more complete characterisation of which redexes can be reduced without breaking Church-Rosserness could be interesting.

@@

We have presented the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.
Though the construction of ranking functions is inherently difficult, using sparse ranking functions, antitone ranking functions and ranking functions w.r.t.~an alternative reduction strategy, we have shown that a great variety of programs can be proved to be AST, including some that are beyond the reach of methods in the literature.

>>>>

Kobayashi et al.~\cite{DBLP:conf/lics/KobayashiLG19} show that the termination probability of order-$n$ probabilistic recursion schemes ($n$-PHORS) can be obtained as the least fixpoint of suitable order-$(n-1)$ fixpoint equations, which are solvable by Kleene fixpoint iteration.  
By contrast, our approach works for programs with continuous distributions. 
Note that $n$-PHORS is strictly less expressive then order-$n$ call-by-name PPCF:
% (because the underlying recursion schemes are not Turing complete):
for example, the PPCF terms in \cref{ex:raven complex,ex:non-affine continuous,ex:higher-order recursion}
%\refExample{complexExample} 
are not definable as PHORS.

@@

The \Cref{def:more general red} of redexes and which positions are acceptable to reduce at is sufficiently restrictive to guarantee confluence, but is also a little more restrictive in some cases than is necessary for this purpose. For example, if the argument of a function is not yet a value, but its reduction to a value would be deterministic, or the function is affine, then applying the function before reducing its argument would not cause any problematic duplication of random samples. Similarly, if $\tsample$ occurs at position $\alpha;@_2;\beta;\lambda;\gamma$, but the function at $\alpha;@_1$ is affine, evaluating the $\tsample$ may in some cases also work fine. A more complete characterisation of which redexes can be reduced without breaking confluence could be interesting.

@@

We have presented the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying a restricted form of confluence.
Though the construction of ranking functions is inherently difficult, using sparse ranking functions, antitone ranking functions and ranking functions w.r.t.~an alternative reduction strategy, we have shown that a great variety of programs can be proved to be AST, including some that are beyond the reach of methods in the literature.

### 2021-04-26 LICS21/supermartingales-proofs.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
For $n \geq 0$, define
\begin{align*}
%M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
\end{align*}
so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \cite{DBLP:conf/icfp/BorgstromLGS16}.
%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
\iffalse
\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
\fi
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed PPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

@@

Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^\omega \subseteq U_i$;
in fact $(U_i)_{\leq l_i} \cdot I^\omega = U_i$.
%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^\omega \subseteq A$.
This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$ 
\begin{equation}
\int_{A \cap U_i}  \mu_{\entrosp}(\dif s) \, g(s) = 
\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
\label{eqn:truncate}
\end{equation}
%(For simplicity, we will write $g' = g$ in the following.)
%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.

Take $s \in U_i$, and set $l = l_i$.
Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
Take $u \in (U_i)_{\leq l}$.
It then follows from the definition of ranking function that
\[%\textstyle
\int_I \Leb(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \leq f(E_i[\widehat{\sigma_i}(u)][\tsample])).
\]
Take $A \in \calF_n$, and integrating both sides, we get
\begin{align*}
& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \\
& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\widehat{\sigma_i}(u)][\tsample])).
\end{align*}
Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
%it follows from the preceding observation that
\begin{align*}
& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\underline{\widehat{\rho}(u')}])) \\
& \leq 
\int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\tsample]))
\end{align*}
and so, by (\ref{eqn:truncate})
\begin{align}
& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])) \nonumber \\
& \leq 
\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])).
\label{eqn:a u ui}
\end{align}

@@

We first show that $T_1$ is bounded i.e.~$T_1 \leq n$, for some $n \in \omega$.

@@

\TMtYStoppingTime*
\begin{proof}
%For each $n \in \omega$, $\set{T_M^\tY = n} \in \calF_{T_n}$.
To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \omega$.
Let $\cal V$ be the set of values in $\Lambda^0$ (which is measurable), and let $\cal W$ be the non-values.
Then, it suffices to observe that $\set{T_M^\tY = n} \cap \set{T_n \leq i}
= \bigcup_{l=n}^i 
\big( M_l^{-1} [{\cal V}] \cap  
M_{l-1}^{-1} [{\cal W}] \cap 
\set{T_n = l} \big)
$, where each of $M_l^{-1} [{\cal V}]$, $M_{l-1}^{-1} [{\cal W}]$, and $\set{T_n = l}$ is in $\calF_i$.
\end{proof}

>>>>

Given an $\mathbb N$-indexed sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
For $n \geq 0$, define
\begin{align*}
%M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
\end{align*}
so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \cite{DBLP:conf/icfp/BorgstromLGS16}.
%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
\iffalse
\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
\fi
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^{\mathbb N}} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed PPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

@@

Observe that if $s \in U_i$ then $\set{s_{\leq l_i}} \cdot I^{\mathbb N} \subseteq U_i$;
in fact $(U_i)_{\leq l_i} \cdot I^{\mathbb N} = U_i$.
%Moreover, for any $A \in \calF_n$, if $s \in A$ then $\set{s_{\leq l}} \cdot I^{\mathbb N} \subseteq A$.
This means that for any measurable $g : \entrosp \to \Real_{\geq 0}$, if $g(s)$ only depends on the prefix of $s$ of length $(l_i+1)$, then, writing $\widehat{g} : I^{l_i+1} \to \Real_{\geq 0}$ where $g(s) = \widehat{g}(s_{\leq l_i+1})$, we have: for any $A \in \calF_n$ 
\begin{equation}
\int_{A \cap U_i}  \mu_{\entrosp}(\dif s) \, g(s) = 
\int_{(A \cap U_i)_{\leq l_i+1}} \Leb_{l_i+1}(\dif t) \, \widehat{g}(t)
\label{eqn:truncate}
\end{equation}
%(For simplicity, we will write $g' = g$ in the following.)
%where $A_{\leq m} := \set{s_{\leq m} \mid s \in A}$.

Take $s \in U_i$, and set $l = l_i$.
Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
Take $u \in (U_i)_{\leq l}$.
It then follows from the definition of ranking function that
\[%\textstyle
\int_I \Leb(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \leq f(E_i[\widehat{\sigma_i}(u)][\tsample])).
\]
Take $A \in \calF_n$, and integrating both sides, we get
\begin{align*}
& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\widehat{\sigma_i}(u)][\underline{r}])) \\
& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\widehat{\sigma_i}(u)][\tsample])).
\end{align*}
Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^{\mathbb N} = U_i$, we have
%it follows from the preceding observation that
\begin{align*}
& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\underline{\widehat{\rho}(u')}])) \\
& \leq 
\int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u') \, f(E_i[\widehat{\sigma_i}(u')][\tsample]))
\end{align*}
and so, by (\ref{eqn:truncate})
\begin{align}
& \int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\underline{\rho(s)}])) \nonumber \\
& \leq 
\int_{A \cap U_i} \mu_{\entrosp}(\dif s) \, f(E_i[\sigma_i(s)][\tsample])).
\label{eqn:a u ui}
\end{align}

@@

We first show that $T_1$ is bounded i.e.~$T_1 \leq n$, for some $n \in \mathbb N$.

@@

\TMtYStoppingTime*
\begin{proof}
%For each $n \in \omega$, $\set{T_M^\tY = n} \in \calF_{T_n}$.
To see $\set{T_M^\tY = n} \in \calF_{T_n}$ for a given $n$, we need to show that $\set{T_M^\tY = n} \cap \set{T_n \leq i} \in \calF_i$, for all $i \in \mathbb N$.
Let $\cal V$ be the set of values in $\Lambda^0$ (which is measurable), and let $\cal W$ be the non-values.
Then, it suffices to observe that $\set{T_M^\tY = n} \cap \set{T_n \leq i}
= \bigcup_{l=n}^i 
\big( M_l^{-1} [{\cal V}] \cap  
M_{l-1}^{-1} [{\cal W}] \cap 
\set{T_n = l} \big)
$, where each of $M_l^{-1} [{\cal V}]$, $M_{l-1}^{-1} [{\cal W}]$, and $\set{T_n = l}$ is in $\calF_i$.
\end{proof}

### 2021-04-26 LICS21/supermartingales.tex
Respond to reviews, "\defn" -> "\emph", "Church-Rosser" -> "confluent"

\begin{definition}\rm
\label{def:ranking function}
A \defn{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}, and
\begin{enumerate}
    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z)$ where $z$ is not free in $N$
    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$ with $R \to R'$
\end{enumerate}
\iffalse
\lo{@Andrew: As defined, ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value. But for AST analysis, we always require $f$ to satisfy this property. It seems cleaner to include this in the definition.} 
\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}
\lo{OK. Let's leave \Cref{def:ranking function} as it is.}
\fi
We say that the ranking function $f$ is \defn{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.

Any closed term for which a ranking (respectively, strict ranking) function exists is called \defn{rankable} (respectively, \defn{strictly rankable}). 
For example, $(\tY \lambda x.x) \, \underline 0$ is not rankable.
\end{definition}

@@

Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$ (i.e.~$\calF_n \subseteq \calF$ is a $\sigma$-algebra, and $\calF_n \subseteq \calF_{n+1}$ for all $n$). 
%(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
Let $T$ be a r.v.~that takes values in $\mathbb{N} \cup \set{\infty}$.
We call $T$ a \defn{stopping time} \emph{adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.

\begin{definition}\rm
\begin{asparaenum}[(i)]
\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \defn{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A \mathbb{P}(\dif \omega) \, Y_{n+1}(\omega) \leq \int_A \mathbb{P}(\dif \omega) \,Y_n(\omega)$).
\item Let $\epsilon > 0$. 
Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
we say that $(Y_n)_{n \geq 0}$ is a $\epsilon$-\defn{ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote. Strange!}\lo{Now it does. Stranger.}
A \defn{ranking supermartingale w.r.t.~$T$} is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
\end{asparaenum}
\end{definition}

@@

Henceforth, fix the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$, and a closed PPCF term $M$.
For $n \geq 0$, define the random variables 
\begin{align*}
M_n(s) &:= \pi_0 (\red^n(M, s))\\
T_M(s) &:= \min \set{n \mid \hbox{$M_n(s)$ is a value}}
\end{align*} 
and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
Thus $T_M$ is the \defn{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \defn{positively almost surely terminating} (\defn{PAST}) if $\expect{T_M} < \infty$.

@@

Let $f$ be a ranking function on $M \in \Lambda^0$.
Define random variables on the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
Y_n(s) & := f(M_{T_n}(s)) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
\end{align}

@@

The random variable $T_M^\tY$, which we call the $\tY$-\defn{runtime} of $M$, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST. 
We say that $M$ is $\tY$-\defn{PAST} if $\expect{T^{\tY}_M} < \infty$.

>>>>

\begin{definition}\rm
\label{def:ranking function}
A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$,\footnote{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.} and
\begin{enumerate}
    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z)$ where $z$ is not free in $N$
    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$ with $R \to R'$
\end{enumerate}
\iffalse
\lo{@Andrew: As defined, ranking function $f$ is not required to satisfy: $f(N) = 0$ iff $N$ is a value. But for AST analysis, we always require $f$ to satisfy this property. It seems cleaner to include this in the definition.} 
\akr{The fact that the ranking function can be 0 in some cases before it reaches a value is necessary to get \Cref{thm:minimal} to work neatly.}
\lo{OK. Let's leave \Cref{def:ranking function} as it is.}
\fi
We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.

Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
For example, $(\tY \lambda x.x) \, \underline 0$ is not rankable.
\end{definition}

@@

Fix a probability space $(\Omega, \calF, \mathbb{P})$ and a filtration $(\calF_n)_{n \geq 0}$ (i.e.~$\calF_n \subseteq \calF$ is a $\sigma$-algebra, and $\calF_n \subseteq \calF_{n+1}$ for all $n$). 
%(In subsequent applications to termination analysis, $\Omega = \entrosp$, and $\mathbb{P} = \mu_{\entrosp}$.)
Let $T$ be a r.v.~that takes values in $\mathbb{N} \cup \set{\infty}$.
We call $T$ a \emph{stopping time adapted to $(\calF_n)_{n \geq 0}$} just if $\set{T = n} \in \calF_n$, for all $n$.

\begin{definition}\rm
\begin{asparaenum}[(i)]
\item A sequence of r.v.s $(Y_n)_{n \geq 0}$ adapted to a filtration $(\calF_n)_{n \geq 0}$ is a \emph{supermartingale} if for all $n \geq 0$, $Y_n$ is integrable (i.e.~$\expect{|Y_n|} < \infty$), and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$ a.s.~(i.e.~for all $A \in \calF_n$, $\int_A Y_{n+1}(\omega) \, \mathbb{P}(\dif \omega) \leq \int_A Y_n(\omega) \, \mathbb{P}(\dif \omega)$).
\item Let $\epsilon > 0$. 
Given a stopping time $T$ and a supermartingale $(Y_n)_{n \geq 0}$, both adapted to filtration $(\calF_n)_{n \geq 0}$, 
we say that $(Y_n)_{n \geq 0}$ is a $\epsilon$-\emph{ranking supermartingale w.r.t.~$T$} if for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.\footnote{For $A \in \calF$, we write ${\bf 1}_A$ for the \emph{indicator function} of $A$, i.e., the random variable defined by ${\bf 1}_A(\omega) := 1$ if $\omega \in A$, and $0$ otherwise.} %\lo{\LaTeX\ does not allow the preceding bracketed sentence to be typeset as a footnote. Strange!}\lo{Now it does. Stranger.}
A \emph{ranking supermartingale w.r.t.~$T$} is a $\epsilon$-ranking supermartingale w.r.t.~$T$ for some $\epsilon > 0$.
\end{asparaenum}
\end{definition}

@@

Henceforth, fix the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$, and a closed PPCF term $M$.
For $n \geq 0$, define the random variables 
\begin{align*}
M_n(s) &:= \pi_0 (\red^n(M, s))\\
T_M(s) &:= \min \set{n \mid \hbox{$M_n(s)$ is a value}}
\end{align*} 
and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
Thus $T_M$ is the \emph{runtime} of $M$ (and $M$ is AST iff $\mu_{\entrosp}(T_M < \infty) = 1$); we say that $M$ is \emph{positively almost surely terminating} (\emph{PAST}) if $\expect{T_M} < \infty$.

@@

Let $f$ be a ranking function on $M \in \Lambda^0$.
Define random variables on the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
Y_n(s) & := f(M_{T_n(s)}(s)) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
\end{align}

@@

The random variable $T_M^\tY$, which we call the $\tY$-\emph{runtime} of $M$, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST. 
We say that $M$ is $\tY$-\emph{PAST} if $\expect{T^{\tY}_M} < \infty$.
